---
title: "Effect of Vitamin C and OJ on Tooth Growth"
author: "F. Alex Crofut"
date: "Saturday, April 23, 2016"
output: pdf_document
---

##Synopsis
This analysis will explore the relationship between tooth growth and dose and supplement, 
both singly and in combination.  First, the analysis will install necessary packages and
the dataset.  Second, the structure of the data will be examined, followed by an initial
exploration of the data.  the assumptions will be stated formally and confidence intervals
will be employed to compare changes in dose and/or supplement type.

##Exploratory Data Analysis

###Required Packages
```{r requiredpackages}
if (!"datasets" %in% installed.packages()) install.packages("datasets")
library("datasets")
if (!"ggplot2" %in% installed.packages()) install.packages("ggplot2")
library("ggplot2")
if (!"gridExtra" %in% installed.packages()) install.packages("gridExtra")
library("gridExtra")
```

###Data Structure
The dataset ToothGrowth (from the datasets package) contains 60 observations on 3 variables:   
```{r data}
data(ToothGrowth)
str(ToothGrowth)
summary(ToothGrowth)
```

### Data Exploration
```{r boxplot}
p1 <- ggplot(aes(x = as.factor(dose), y = len), data = ToothGrowth) + 
        geom_boxplot(fill = "blue") +  
        labs(x = "Dose", y = "Tooth Growth") + 
        ggtitle("Tooth Growth by Dose")
p2 <- ggplot(aes(x = supp, y = len), data = ToothGrowth) + 
        geom_boxplot(aes(fill = supp)) + facet_wrap(~ dose) + 
        labs(x = "Supplement", y = "Tooth Growth") + 
        ggtitle("Tooth Growth by Dose and Supplement")
grid.arrange(p1, p2, ncol=2)
```
From the first boxplot, it appears that dose is indeed correlated with tooth growth.  Furthermore,
the second boxplot appears to show that the supplement choice has an effect at the two lower doses, 
but not at a dose of 2.0mg.

##Key Assumptions
In order to perform the following hypothesis tests, the following assumptions need to be made:   

* Population is independent an didentically distributed
* Data is not paired
* Measurement error is accounted for by significant digits
* The study was double blind
* A 95% confidence is sufficient to determine significance in this case

##Hypothesis Tests

###Dose as only Determinant
```{r dose only}
d1 <- subset(ToothGrowth, dose %in% c(0.5, 1.0))
d2 <- subset(ToothGrowth, dose %in% c(0.5, 2.0))
d3 <- subset(ToothGrowth, dose %in% c(1.0, 2.0))
t1 <- t.test(len ~ dose, paired=FALSE, var.equal = FALSE, data = d1)
t2 <- t.test(len ~ dose, paired=FALSE, var.equal = FALSE, data = d2)
t3 <- t.test(len ~ dose, paired=FALSE, var.equal = FALSE, data = d3)
```
The confidence intervals:   

* [`r round(t1$conf.int,3)`] for increasing dose from 0.5mg to 1.0mg   
* [`r round(t2$conf.int,3)`] for increasing dose from 0.5mg to 2.0mg   
* [`r round(t3$conf.int,3)`] for increasing dose from 1.0mg to 2.0mg   

All allow rejection of the null hypothesis that dose has no effect on tooth growth.

###Supplement as only Determinant
```{r supplement only}
supp <- t.test(len ~ supp, paired = FALSE, var.equal = FALSE, data = ToothGrowth)
```
The confidence interval [`r round(supp$conf.int, 3)`] means the null hypothesis that 
Tooth Length is not affected by supplement choice **cannot** be rejected.

###Supplement Type as a Factor within Dose Levels
```{r both}
d4 <- subset(ToothGrowth, dose == 0.5)
d5 <- subset(ToothGrowth, dose == 1.0)
d6 <- subset(ToothGrowth, dose == 2.0)
t4 <- t.test(len ~ supp, paired = FALSE, var.equal = FALSE, data = d4)
t5 <- t.test(len ~ supp, paired = FALSE, var.equal = FALSE, data = d5)
t6 <- t.test(len ~ supp, paired = FALSE, var.equal = FALSE, data = d6)
```
The confidence intervals for increasing the dose from 0.5mg to 1.0mg [`r round(t4$conf.int,3)`] 
and increasing the dose from 0.5mg to 2.0mg [`r round(t5$conf.int,3)`] provide evidence that 
the OJ is more effective than Vitaming C in increasing tooth length.  However, the interval 
for increasing the dose from 1.0mg to 2.0mg [`r round(t6$conf.int,3)`] shows that the choice 
of supplement does not affect the effectiveness of the dose at 2.0mg.

##Conclusions
Given the key assumptions and the results of the t-tests, one can infer that increases in dose
have a significant correlation to tooth growth, regardless of the choice in supplements.  Furthermore,
the choice of supplements impacts the effectiveness of doses 0.5mg and 1.0mg.  However, for doses
of 2.0mg, the choice of supplement is no longer a significant effect.  Finally, there is no evidence
that the choice of supplement alone has a significant correlation with tooth growth.